Eco-evolutionary dynamics of finite populations from demographic first principles

### 2023-01-24 ind_files/03_BD_processes.tex
Add minor footnote in chap 4

\epigraph{\justifying ``What we want to know is how the biological forces of natality and mortality are so integrated and correlated in their action as to lead to a final result in size of population which follows this particular curve rather than some other one"}{Raymond Pearl}

Raymond Pearl was one of the pioneers of mathematical ecology as a discipline, and the quote in the epigraph is from his 1927 book titled \textit{The Growth of Populations}. Pearl realized 95 years ago that population dynamics must be ultimately explained by the mechanistic processes of birth and death. Today, there are mounting calls for more mechanistic models of evolution that are grounded in these fundamental processes of birth and death \citep{geritz_mathematical_2012,doebeli_towards_2017}. Ecologists also increasingly recognize that incorporating stochasticity is vital to developing more realistic ecological models \citep{hastings_transients_2004, coulson_skeletons_2004, boettiger_noise_2018, shoemaker_integrating_2020,schreiber_does_2022} and does more than `add noise' to deterministic expectations. Individual-based models, where ecological rules are specified at the level of the individual, are a powerful mathematical tool for mechanistic descriptions of stochastic population dynamics \citep{black_stochastic_2012}. Birth-death processes are a very general class of stochastic processes that can be used to capture a wide range of eco-evolutionary processes. `System-size expansions' and their subsequent analysis using the `weak noise approximation' are common tools for analyzing stochastic birth-death processes that are well-known in the statistical physics and applied mathematics communities \citep{gardiner_stochastic_2009}. However, these tools remain relatively underappreciated in ecology, despite being relatively easy to understand and extremely well-motivated in scenarios germane to ecology and evolution. Here, I present a formulation of population dynamics constructed from first principles grounded in birth-death processes. To facilitate readership by a broad audience, only passing familiarity with calculus (derivatives, integrals, Taylor expansions) and probability are assumed. Familiarity with stochastic calculus is helpful for some sections but is not required, and I provide a brief introduction below. I begin by introducing birth-death processes, SDEs, and the Fokker-Planck equation in a non-technical manner, and providing the intuition for system-size expansions and weak noise analysis in ecological systems. In section \ref{sec_1D_processes}, I show how fluctuating population size of populations of identical individuals can be tracked through a one-dimensional birth-death process. I introduce a description of the system via a `master equation', and then conduct a `system-size expansion' to obtain a Fokker-Planck equation for the system. Finally, I conduct a weak noise approximation to arrive at a linear Fokker-Planck equation which can be solved exactly using some stochastic calculus to arrive at a closed-form solution given by a time-dependent Ornstein-Uhlenbeck process. As an example, I illustrate the complete process for a stochastic version of the logistic equation. In section \ref{sec_nD_processes}, I present a multivariate process to describe the evolution of discretely varying traits, and, as before, illustrate the system size expansion and the weak noise approximation. Under mild assumptions, we show that the deterministic limit of this process is the well-known replicator-mutator equation (or equivalently, Eigen's quasispecies equation), thus establishing the microscopic basis of well-known equations in evolutionary game theory. I also show that the mean value of the trait in the population changes according to the Price equation in the deterministic limit. Section \ref{sec_infD_processes} introduces a function-valued process to model the evolution of quantitative traits such as body size, which can take on uncountably many values. This function-valued process can then also be analyzed via a system-size approximation to arrive at a `functional' Fokker-Planck equation, in terms of functional derivatives. Under mild assumptions, we show that classic equations such as Kimura's infinite alleles model and the canonical equation of adaptive dynamics can be derived as the deterministic limits of this stochastic process. We also conduct a weak noise approximation to arrive at a linear functional Fokker-Planck equation and illustrate the process using the stochastic quantitative logistic equation introduced by \citep{doebeli_adaptive_2011} and \citep{rogers_demographic_2012} as an example.  Sections \ref{sec_1D_processes} and \ref{sec_nD_processes} provide a unified, pedagogical review and conceptual synthesis of techniques that are well-known in physics, as applied to population biology. Section \ref{sec_infD_processes}, to the best of my knowledge, is entirely original. Section \ref{sec_infD_processes} also presents a simple heuristic derivation of quantitative genetics models and adaptive dynamics from stochastic first principles that is simpler than the rigorous mathematical derivations grounded in measure theory and martingale/markov theory that are currently standard reference in theoretical ecology \citep{champagnat_individual_2008}.

@@

We will evaluate the integrals on the RHS of \eqref{nD_intermediate_for_moments} using integration by parts. Recall that the general formula for integration by parts of two functions $u$ and $v$ defined on a domain $\Omega$ is given by:

@@

where $\partial \Omega$ is the boundary of $\Omega$, $dS$ is the surface element of this boundary, and $\gamma$ is the unit outward normal to the boundary. In our case, we have $\Omega = \mathbb{R}^m$, and thus the boundary conditions are evaluated as $\|y\| \to \infty$. We assume that stochastic fluctuations do not grow too large, and therefore impose the condition $\displaystyle \lim_{\|y\| \to \infty}  P(\mathbf{y},t) = 0$. Further, we assume that this decay is fast enough that $\displaystyle \lim_{\|y\| \to \infty}D_jP(\mathbf{y},t) = 0\ \forall \ j$. Under these conditions, we can evaluate the two integrals in the RHS of \eqref{nD_intermediate_for_moments} by using integration by parts and discarding the boundary term (The second term on the RHS of \eqref{int_by_parts_general_formula}). Note that since the $y_i$s are orthogonal to each other, we have the relation:

@@

\subsection{An example: The quantitative stochastic logistic equation}
As an example, let us carry out the functional Kramers-Moyal expansion for the birth and death functionals given by \eqref{Rogers_logistic_BD}. In terms of the scaled variable $\phi = K\nu$, these functions read:
\begin{equation}
\label{Rogers_logistic_BD_scaled}
\begin{aligned}
b_K(x|\phi) &= \frac{1}{K}b(x|\nu) = \frac{1}{K}\left( r\int\limits_{\mathcal{T}}m(x,y)K\phi(y)dy\right)\\
    d_K(x|\phi) &= \frac{1}{K}d(x|\nu) =  \frac{1}{K}\left(\frac{K\phi(x)}{Kn(x)}\int\limits_{\mathcal{T}}\alpha(x,y)K\phi(y)dy\right)
\end{aligned}
\end{equation}
Thus, using equation \eqref{deterministic_traj}, the deterministic trajectory becomes:
\begin{equation}
\label{Rogers_logistic_BD_deterministic}
\frac{\partial \psi}{\partial t}(x,t) = r\int\limits_{\mathcal{T}}m(x,y)\psi(y,t)dy-\frac{1}{n(x)}\psi(x,t)\int\limits_{\mathcal{T}}\alpha(x,y)\psi(y,t)dy
\end{equation}
Note that if we employ the change of variables $\Psi = K\psi$ to go back from $\mathcal{M}_{K}$ (\textit{i.e} $\phi^{(t)}$) to $\mathcal{M}$ (\textit{i.e} $\nu^{(t)}$), we recover the familiar continuous logistic equation (Equation \eqref{cts_logistic}) as the deterministic limit:
\begin{align*}
\frac{\partial \Psi}{\partial t}(x,t) &= r\int\limits_{\mathcal{T}}m(x,y)\Psi(y,t)dy-\frac{\Psi(x,t)}{Kn(x)}\int\limits_{\mathcal{T}}\alpha(x,y)\Psi(y,t)dy \\
&\approx r\Psi(x,t) -\frac{\Psi(x,t)}{K(x)}\int\limits_{\mathcal{T}}\alpha(x,y)\Psi(y,t)dy + D_m\nabla^2_{x}\Psi(x,t)
\end{align*}
where $K(x) = Kn(x)$ is the carrying capacity experienced by an individual of phenotype $x$, and $D_m = r \sigma_m^2/2$ measures the `diffusion rate' of the population in trait space. It is left as an exercise for the reader to verify by the same steps that if we instead have the birth rate functional $b(x|\phi) = \lambda\int m(x,y)\phi(y)dy$ (with $m(x,y)$ as defined in \eqref{Rogers_logistic_BD}) and the death rate functional $d(x|\phi) = \phi(x)\left(\mu+(\lambda-\mu)\phi(x)/K\right)$ corresponding to \emph{linear} density-dependent death, the macroscopic limit yields the famous Fisher-KPP equation with growth rate $r=\lambda-\mu$ and diffusion constant $D = \lambda \sigma_m^2/2$.
% In any case, for the system defined by \eqref{Rogers_logistic_BD_scaled}, we can also calculate $\mathcal{D}_{\zeta}[\mathcal{A}^-]$ as
% \begin{align*}
% \mathcal{D}_{\zeta}[\mathcal{A}^-] &= \frac{d}{d\epsilon}\left( r\int\limits_{\mathcal{T}} m(x,y)(\psi(y)+\epsilon\zeta(y))dy - \frac{\psi(x)+\epsilon\zeta(x)}{n(x)}\int\limits_{\mathcal{T}}\alpha(x,y)(\psi(y)+\epsilon\zeta(y))dy\right) \biggl{|}_{\epsilon = 0}\\
% &= r\int\limits_{\mathcal{T}}m(x,y)\zeta(y)dy - \frac{1}{n(x)}\left(\psi(x)\int\limits_{\mathcal{T}}\alpha(x,y)\zeta(y)dy + \zeta(x)\int\limits_{\mathcal{T}}\alpha(x,y)\psi(y)dy\right)
% \end{align*}

>>>>

\epigraph{\justifying Somewhere [ ... ] between the specific that has no meaning and the general that has no content there must be, for each purpose and at each level of abstraction, an optimum degree of generality}{Kenneth Boulding}

Raymond Pearl, one of the pioneers of mathematical ecology as a discipline, wrote in 1927 that ``What we want to know is how the biological forces of natality and mortality are so integrated and correlated in their action as to lead to a final result in size of population which follows this particular curve rather than some other one'' \citep{pearl_growth_1976}. Pearl realized 95 years ago that population dynamics must be ultimately explained by the mechanistic processes of birth and death. Today, there are mounting calls for more mechanistic models of evolution that are grounded in these fundamental processes of birth and death \citep{geritz_mathematical_2012,doebeli_towards_2017}. Ecologists also increasingly recognize that incorporating stochasticity is vital to developing more realistic ecological models \citep{hastings_transients_2004, coulson_skeletons_2004, boettiger_noise_2018, shoemaker_integrating_2020,schreiber_does_2022} and does more than `add noise' to deterministic expectations. Individual-based models, where ecological rules are specified at the level of the individual, are a powerful mathematical tool for mechanistic descriptions of stochastic population dynamics \citep{black_stochastic_2012}. Birth-death processes are a very general class of stochastic processes that can be used to capture a wide range of eco-evolutionary processes. `System-size expansions' and their subsequent analysis using the `weak noise approximation' are common tools for analyzing stochastic birth-death processes that are well-known in the statistical physics and applied mathematics communities \citep{gardiner_stochastic_2009}. However, these tools remain relatively underappreciated in ecology, despite being relatively easy to understand and extremely well-motivated in scenarios germane to ecology and evolution. Here, I present a formulation of population dynamics constructed from first principles grounded in birth-death processes. To facilitate readership by a broad audience, only passing familiarity with calculus (derivatives, integrals, Taylor expansions) and probability are assumed. Familiarity with stochastic calculus is helpful for some sections but is not required, and I provide a brief introduction below. I begin by introducing birth-death processes, SDEs, and the Fokker-Planck equation in a non-technical manner, and providing the intuition for system-size expansions and weak noise analysis in ecological systems. In section \ref{sec_1D_processes}, I show how fluctuating population size of populations of identical individuals can be tracked through a one-dimensional birth-death process. I introduce a description of the system via a `master equation', and then conduct a `system-size expansion' to obtain a Fokker-Planck equation for the system. Finally, I conduct a weak noise approximation to arrive at a linear Fokker-Planck equation which can be solved exactly using some stochastic calculus to arrive at a closed-form solution given by a time-dependent Ornstein-Uhlenbeck process. As an example, I illustrate the complete process for a stochastic version of the logistic equation. In section \ref{sec_nD_processes}, I present a multivariate process to describe the evolution of discretely varying traits, and, as before, illustrate the system size expansion and the weak noise approximation. Under mild assumptions, we show that the deterministic limit of this process is the well-known replicator-mutator equation (or equivalently, Eigen's quasispecies equation), thus establishing the microscopic basis of well-known equations in evolutionary game theory. I also show that the mean value of the trait in the population changes according to the Price equation in the deterministic limit. Section \ref{sec_infD_processes} introduces a function-valued process to model the evolution of quantitative traits such as body size, which can take on uncountably many values. This function-valued process can then also be analyzed via a system-size approximation to arrive at a `functional' Fokker-Planck equation, in terms of functional derivatives. Under mild assumptions, we show that classic equations such as Kimura's infinite alleles model and the canonical equation of adaptive dynamics can be derived as the deterministic limits of this stochastic process. We also conduct a weak noise approximation to arrive at a linear functional Fokker-Planck equation and illustrate the process using the stochastic quantitative logistic equation introduced by \citep{doebeli_adaptive_2011} and \citep{rogers_demographic_2012} as an example.  Sections \ref{sec_1D_processes} and \ref{sec_nD_processes} provide a unified, pedagogical review and conceptual synthesis of techniques that are well-known in physics, as applied to population biology. Section \ref{sec_infD_processes}, to the best of my knowledge, is entirely original. Section \ref{sec_infD_processes} also presents a simple heuristic derivation of quantitative genetics models and adaptive dynamics from stochastic first principles that is simpler than the rigorous mathematical derivations grounded in measure theory and martingale/markov theory that are currently standard reference in theoretical ecology \citep{champagnat_individual_2008}.

@@

We will evaluate the integrals on the RHS of \eqref{nD_intermediate_for_moments} using integration by parts. Recall that for any two functions $u$ and $v$ defined on a domain $\Omega$, the general formula for integration by parts is given by:

@@

where $\partial \Omega$ is the boundary of $\Omega$, $dS$ is the surface element of this boundary, and $\gamma_i$ is the $i\textsuperscript{th}$ component of the unit outward normal to the boundary. In our case, we have $\Omega = \mathbb{R}^m$, and thus the boundary conditions are evaluated as $\|y\| \to \infty$. We assume that the magnitude of stochastic fluctuations is bounded, and therefore impose the condition $\displaystyle \lim_{\|y\| \to \infty}  P(\mathbf{y},t) = 0$. Further, we assume that this decay is fast enough that $\displaystyle \lim_{\|y\| \to \infty}D_jP(\mathbf{y},t) = 0\ \forall \ j$. Under these conditions, we can evaluate the two integrals in the RHS of \eqref{nD_intermediate_for_moments} by using integration by parts and discarding the boundary term (The second term on the RHS of \eqref{int_by_parts_general_formula}). Note that since the $y_i$s are orthogonal to each other, we have the relation:

@@

### 2023-01-24 ind_files/04_polymorphisms.tex
Add minor footnote in chap 4

which is the familiar version of the replicator equation seen in most textbooks, with an extra $N_K(t)$ factor to account for the fact that $\sum_i x_i$ is allowed to fluctuate in our model. If instead $N_K$ was a constant for all time, it could simply be absorbed into the definition of the payoff matrix $B$ to obtain exactly the replicator equation as presented in most ecology/evolution textbooks.\\

@@

% or, in matrix form:
% \begin{equation}
% \resizebox{0.91\textwidth}{!}{$\displaystyle
% \frac{d}{dt}\begin{bmatrix}
%     \mathbb{E}[y_1]\\
%     \mathbb{E}[y_2]\\
%     \vdots\\
%     \mathbb{E}[y_i]\\
%     \vdots\\
%     \mathbb{E}[y_m]
%     \end{bmatrix}
%     =
% \begin{bmatrix}
%     (r + \sum\limits_{j=1}^{m}B_{1j}a_j + a_1B_{11}) & a_1B_{12} & a_1B_{13} & \dots & \dots & \dots & a_1B_{1m}\\
%     a_2B_{21}) & (r + \sum\limits_{j=1}^{m}B_{2j}a_j + a_2B_{22}) & a_2B_{23} & \dots & \dots & \dots & a_2B_{2m}\\
%     \vdots & \vdots & \vdots &\vdots & \vdots & \vdots & \vdots\\
%     a_{i}B_{i1} & a_iB_{i2} & a_iB_{i3} & \dots & (r + \sum\limits_{j=1}^{m}B_{ij}a_j + a_iB_{ii}) & \dots & a_iB_{im}\\
%     \vdots & \vdots & \vdots &\vdots & \vdots & \vdots & \vdots\\
%     a_mB_{m1} & a_mB_{m2} & a_mB_{m3} & \dots & \dots & \dots & (r + \sum\limits_{j=1}^{m}B_{mj}a_j + a_mB_{mm})
%     \end{bmatrix}
% \begin{bmatrix}
%     \mathbb{E}[y_1]\\
%     \mathbb{E}[y_2]\\
%     \vdots\\
%     \mathbb{E}[y_i]\\
%     \vdots\\
%     \mathbb{E}[y_m]
%     \end{bmatrix}
% $}
% \end{equation}

@@

In section \ref{sec_infD_processes}, we used various approximations to arrive at the linear functional Fokker-Planck equation
\begin{equation}
\label{functional_WNE_chap_4}
    \frac{\partial P}{\partial t}(\zeta,t) = \int\limits_{\mathcal{T}}\left(-\frac{\delta}{\delta \zeta(x)}\left\{\mathcal{D}_{\zeta}[\mathcal{A}^{-}](x)P(\zeta,t)\right\}+\frac{1}{2}\mathcal{A}^{+}(x|\psi)\frac{\delta^2}{\delta\zeta(x)^2}\{P(\zeta,t)\}\right)dx
\end{equation}
for describing stochastic fluctuations $\zeta$ from the deterministic solution obtained by solving \eqref{deterministic_traj}. Our goal is now to find a method to effectively detect and describe evolutionary branches (modes in trait space, corresponding to individual morphs) for this process. We will do this by measuring the autocorrelation of the distribution of the population over trait space. A convenient theorem due to Weiner and Khinchin relates the autocorrelation of a probability distribution to its power spectral density via Fourier transformation. This has been extensively used in spatial ecology, and we too will make use of it here. Specifically, we will carry out a basis expansion of our functions in the Fourier basis $\{e^{ikx}\}_{k\in\mathbb{Z}}$. If $\mathcal{D}_{\zeta}[\mathcal{A}^{-}]$ is a translation-invariant linear operator, then $\exp(ikx)$ acts as an eigenfunction, significantly simplifying the calculations. We therefore assume this below.

@@

The presence of phenotypic clustering can be detected using the `spatial covariance' of our original process $\phi$, defined as \cite{rogers_demographic_2012}:
\begin{equation}
\label{spatial_covariance_defn}
\Xi[x] = m(\mathcal{T})\int\limits_{\mathcal{T}}\mathbb{E}[\phi_{\infty}(x)\phi_{\infty}(y-x)]dy
\end{equation}
where $\phi_{\infty}$ is the stationary state distribution of $\{\phi_t\}_{t}$ and $m$ is the Lebesgue measure. We can use a spatial analogue of the Wiener-Khinchin theorem to calculate:
\begin{equation}
\label{spatial_covariance_zeta}
\Xi[x] = m(\mathcal{T})\left[\int\limits_{\mathcal{T}}\psi_{\infty}(x)\psi_{\infty}(y-x)dy + \frac{1}{K}\sum\limits_{r=-\infty}^{\infty}\mathbb{E}[|\zeta_r|^2]e^{irx}\right]
\end{equation}
where the expectations in the second term are for the stationary state. A flat $\Xi[x]$ indicates that there are no clusters, and peaks indicate the presence of clusters.

>>>>

which is the familiar version of the replicator equation seen in most textbooks, with an extra $N_K(t)$ factor to account for the fact that $\sum_i x_i$ is allowed to fluctuate in our model. If instead $N_K$ was a constant for all time, it could simply be absorbed into the definition of the payoff matrix $B$ to obtain exactly the replicator equation as presented in most ecology/evolution textbooks. We briefly note here that if $m$ is large, then both the stochastic dynamics \eqref{nD_example_SDE} and the deterministic limit \eqref{nD_example_det_limit} can be simplified from an $m$ dimensional system to an $m-1$ dimensional system by a coordinate transformation. If we go from the variables $x_1,\ldots,x_m$ to the variables $p_1,\ldots,p_{m-1},N_K$, then, we can exploit the fact that $N_K$ varies much less than the $p_i$ terms to project the system onto a `slow manifold' in which $N_K$ is approximately constant, thus reducing to an $m-1$ dimensional system of equations and obtaining the relation between the Lotka-Volterra equations for $m$ species and the replicator equation for $m-1$ tactics. However, we will not explore this further in this manuscript, and refer the reader to \citep{constable_stochastic_2013} for a review of the ideas of dynamics on slow manifolds.\\

@@

or, in matrix form:
\resizebox{0.93\textwidth}{!}{$\displaystyle
	\frac{d}{dt}\begin{bmatrix}
	\mathbb{E}[y_1]\\
	\mathbb{E}[y_2]\\
	\vdots\\
	\mathbb{E}[y_i]\\
	\vdots\\
	\mathbb{E}[y_m]
	\end{bmatrix}
	=
	\begin{bmatrix}
	(r + \sum\limits_{j=1}^{m}B_{1j}a_j + a_1B_{11}) & a_1B_{12} & a_1B_{13} & \dots & \dots & \dots & a_1B_{1m}\\
	a_2B_{21} & (r + \sum\limits_{j=1}^{m}B_{2j}a_j + a_2B_{22}) & a_2B_{23} & \dots & \dots & \dots & a_2B_{2m}\\
	\vdots &  & \ddots & &  & & \vdots\\
	a_{i}B_{i1} & a_iB_{i2} & a_iB_{i3} & \dots & (r + \sum\limits_{j=1}^{m}B_{ij}a_j + a_iB_{ii}) & \dots & a_iB_{im}\\
	\vdots &  &  & & & \ddots & \vdots\\
	a_mB_{m1} & a_mB_{m2} & a_mB_{m3} & \dots & \dots & \dots & (r + \sum\limits_{j=1}^{m}B_{mj}a_j + a_mB_{mm})
	\end{bmatrix}
	\begin{bmatrix}
	\mathbb{E}[y_1]\\
	\mathbb{E}[y_2]\\
	\vdots\\
	\mathbb{E}[y_i]\\
	\vdots\\
	\mathbb{E}[y_m]
	\end{bmatrix}
	$}
The eigenvalues of the first matrix on the RHS will tell us whether the fixed point $\mathbb{E}[y_i] = 0 \ \forall \ i$ (the only fixed point of this system) is stable, or whether fluctuations are expected to grow (up to the point where the fluctuations are so large that the WNA is no longer valid).

@@

In section \ref{sec_infD_processes}, we used various approximations to arrive at the linear functional Fokker-Planck equation
\begin{equation}
\label{functional_WNE_chap_4}
    \frac{\partial P}{\partial t}(\zeta,t) = \int\limits_{\mathcal{T}}\left(-\frac{\delta}{\delta \zeta(x)}\left\{\mathcal{D}_{\zeta}[\mathcal{A}^{-}](x)P(\zeta,t)\right\}+\frac{1}{2}\mathcal{A}^{+}(x|\psi)\frac{\delta^2}{\delta\zeta(x)^2}\{P(\zeta,t)\}\right)dx
\end{equation}
for describing stochastic fluctuations $\zeta$ from the deterministic solution obtained by solving \eqref{deterministic_traj}. Our goal is now to find a method to effectively detect and describe evolutionary branches (modes in trait space, corresponding to individual morphs) for this process. We will do this by measuring the autocorrelation of the distribution of the population over trait space. A convenient theorem due to Weiner and Khinchin relates the autocorrelation of a probability distribution to its power spectral density via Fourier transformation. This has been extensively used in spatial ecology, and we too will make use of it here. Specifically, we will carry out a basis expansion of our functions in the Fourier basis $\{e^{ikx}\}_{k\in\mathbb{Z}}$. If $\mathcal{D}_{\zeta}[\mathcal{A}^{-}]$ is a translation-invariant\footnote{This is horrible nomenclature by the mathematicians. Though `invariant' is the conventional name for this concept, the intended meaning is not really invariant but `equivariant'. Formally, let $\mathcal{F}$ be a suitable function space of real valued functions. For any $c \in \mathbb{R}$, let $T_c: \mathcal{F} \to \mathcal{F}$ be the translation operator on this space, defined by $T_c[f(x)] = f(x+c)$. An operator $L: \mathcal{F} \to \mathcal{F}$ is said to be translation-invariant if it commutes with $T_c$ for every $c \in \mathbb{R}$, \emph{i.e.} $T_c[L[f]] = L[T_c[f]] \ \forall \ f \in \mathcal{F} \ \forall \ c \in \mathbb{R}$} linear operator, then $\exp(ikx)$ acts as an eigenfunction, significantly simplifying the calculations. We therefore assume this below.

@@

The presence of phenotypic clustering can be detected using the `spatial covariance' of our original process $\phi$, defined as \citep{rogers_demographic_2012}:
\begin{equation}
\label{spatial_covariance_defn}
\Xi[x] = m(\mathcal{T})\int\limits_{\mathcal{T}}\mathbb{E}[\phi_{\infty}(x)\phi_{\infty}(y-x)]dy
\end{equation}
where $\phi_{\infty}$ is the stationary state distribution of $\{\phi_t\}_{t}$ and $m$ is the Lebesgue measure. We can use a spatial analogue of the Wiener-Khinchin theorem to calculate:
\begin{equation}
\label{spatial_covariance_zeta}
\Xi[x] = m(\mathcal{T})\left[\int\limits_{\mathcal{T}}\psi_{\infty}(x)\psi_{\infty}(y-x)dy + \frac{1}{K}\sum\limits_{r=-\infty}^{\infty}\mathbb{E}[|\zeta_r|^2]e^{irx}\right]
\end{equation}
where the expectations in the second term are for the stationary state. A flat $\Xi[x]$ indicates that there are no clusters, and peaks indicate the presence of clusters.
